Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-05-15 src/tex/ms.tex
update figures

%\definecolor{rb4}{HTML}{27408B}
%\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
%\newcommand{\kl}[1]{\color{cyan}[KL: #1 ]}
\newcommand{\ripple}{\texttt{ripple}}
\newcommand{\python}{\texttt{python}}
\newcommand{\jax}{\texttt{jax}}
\newcommand{\zdethp}{\texttt{zdethp}}
\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}

@@

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{The value of loss function against number of iterations. The code is terminated after 30000 steps }
	\label{fig:loss}
\end{figure}

@@

\begin{figure}[t]
	\script{intrin_space.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/intrin_space.pdf}
	\caption{Parameter space with mass ratio against normalized reduced spin. Orange: Training waveforms; Blue: Testing waveforms}
	\label{fig:intrin_space}
\end{figure}

@@

To quantify the effect of optimization over the entire dataset, We evaluate the
mismatch of all testing waveforms using a constant PSD weighted loss function,
$\mathcal{L}_{\mathrm{mean}}$. We present the resulting distribution in Fig.
\ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{fl}}$, we observe a similar
improvement with a 22.9\% decrease in the median. Even though the distribution
lacks a clear peak due to the problem discussed in Section
\ref{subsec:optimization}, both distributions show observable improvement, and
the optimization scheme remains unaffected \kw{What about the optimization scheme?}.

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of waveform mismatches calculated using both
	$\mathcal{L}_{\mathrm{mean}}$ and $\mathcal{L}_{\mathrm{fl}}$. We use 
	training waveforms listed in Tab.~\ref{tab:q148} and
	mismatches are weighted with the constant PSD. For the
	$\mathcal{L}_{\mathrm{mean}}$ distributions, the median (dashed lines) decreased by 50.0\%
	while the median of $\mathcal{L}_{\mathrm{fl}}$ distribution decreased by
	22.9\%.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The
	distribution labeled \texttt{q148} uses training waveforms listed in
	Tab.~\ref{tab:q148} while the \texttt{q1248} distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using the constant PSD with the loss function $\mathcal{L}_{\mathrm{mean}}$.}
	\label{fig:q148_q1248_compare}
\end{figure}

\kw{Which plot is this paragraph referring to?}
Applying the same methods, we observe that the distributions of mismatches
calculated using the {\zdethp} PSD exhibit superior improvement compared to the
weighted mismatch \kw{What do you mean by weighted mismatch? Like flat PSD?}.
The shape of the distribution is similar to that shown in Fig. \ref{fig:q148}.
This result is expected, as the IMRPhenomD model was initially constructed and
fitted using the {\zdethp} weighted mismatch. Consequently, the model is
anticipated to closely fit the NR waveforms while incorporating the influence of
the {\zdethp} PSD, rather than a constant PSD.

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248} \kw{What
special about the q=2 events?}. The new set of coefficients generated from this
optimization process yields only marginal improvements in the newly produced
waveforms, as seen in Fig. \ref{fig:q148_q1248_compare}. The high mismatch tail
of the optimized distribution remains comparable in length and endpoint to the
original distribution, indicating that our procedure is incapable of improving
these waveforms. Similarly, utilizing the {\zdethp} PSD to optimize the loss
function with additional waveforms leads to similar improvement in the resulting
distribution.

@@

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ 
and $\chi_2<0$) regions respond to optimization differently. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right 
region do not improve after optimization. Most waveforms even turned worse after 
optimization.

\begin{figure}[t]
	\script{ps_q148_qchi.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_qchi.pdf}
	\caption{Parameter space of testing waveforms of
	$q$~vs.~$\chi_{\mathrm{PN}}$. We use the recalibrated result from
	$\mathcal{L}_{\mathrm{mean}}$ with the constant noise spectrum and training
	waveforms in Tab.~\ref{tab:q148}. Here, the colorbar represents the
	$\log_{10}$ difference between optimized and original unweighted
	mismatches.}
	\label{fig:ps_q148_qchi}
\end{figure}
\begin{figure}[t]
	\script{ps_q148_chi1chi2.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_chi1chi2.pdf}
	\caption{Parameter space of testing waveforms of $\chi_1$~vs.~$\chi_2$. We
	use the recalibrated result from $\mathcal{L}_{\mathrm{mean}}$ with the
	constant noise spectrum and training waveforms in Tab.~\ref{tab:q148}.}
	\label{fig:ps_q148}
\end{figure}

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0167 & 8.0 & 0.00     & 0.00     \\ \midrule SXS:BBH:0370 & 1.0
		& -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule SXS:BBH:0156 & 1.0
		& -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule SXS:BBH:0304 & 1.0
		& 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 regions. From top to down are the top-right ($\chi_1,\chi_2>0$), top-left ($\chi_1<0<\chi_2$), bottom-left ($\chi_1,\chi_2<0$) and bottom-right ($\chi_1>0>\chi_2$) regions. Note that for the top-right and bottom-left regions, waveforms are chosen to have equal or similar spins, while the training waveforms for the other two regions are chosen to have opposite spins.}
	\label{tab:quadrants}
\end{table}

@@

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left and bottom-right regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ for the loss function. Waveforms in the top-left region generally improves while waveforms in the bottom-right region worsened, as indicated by the median (dashed lines). \kl{I tried to overlap these two histograms together but it makes them very hard to read. }}
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently. Colorbar
	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

In this paper, we have presented a systematic method to recalibrate GW models.
This method utilizes {\jax}'s automatic differentiation to apply
derivative-based optimization to recalibrate GW models jointly. Using the new
implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
mismatches have a slightly better improvement. In contrast, different types of
loss function result in significantly different final mismatch distributions. As 
seen in Fig.~\ref{fig:q148}, $\mathcal{L}_{\mathrm{mean}}$ outperforms 
$\mathcal{L}_{fl}$. By increasing the number of training waveforms, we see a 
slight improvement increase in Fig.~\ref{fig:q148_q1248_compare}.

>>>>

\newcommand{\ripple}{\texttt{ripple}}
\newcommand{\python}{\texttt{python}}
\newcommand{\jax}{\texttt{jax}}
\newcommand{\zdethp}{\texttt{zdethp}}
\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}

@@

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{Loss functions against number of iterations. We take the set of coefficients at the minimum of the validation loss.}
	\label{fig:loss}
\end{figure}

@@

\begin{figure}[t]
	\script{intrin_space.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/intrin_space.pdf}
	\caption{Parameter space with mass ratio $q$ against post-newtonian spin parameter $\chi_{\mathrm{PN}}$. Orange: Training waveforms; Blue: Testing waveforms}
	\label{fig:intrin_space}
\end{figure}

@@

To quantify the effect of optimization over the entire dataset, We evaluate the
mismatch of all testing waveforms using a constant PSD weighted loss function,
$\mathcal{L}_{\mathrm{mean}}$. We present the resulting distribution in Fig.
\ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{fl}}$, we observe a similar
improvement with a 22.9\% decrease in the median. Even though the distribution
lacks a clear peak due to the problem discussed in Section
\ref{subsec:optimization}, both distributions show observable improvement.

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of waveform mismatches calculated using 
	$\mathcal{L}_{\mathrm{mean}}$. We use training waveforms listed in 
	Tab.~\ref{tab:q148} and	mismatches are weighted with the constant PSD. 
	Dashed lines represent the median of the distributions, which has decreased by 45.3\%.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The
	distribution labeled \texttt{q148} uses training waveforms listed in
	Tab.~\ref{tab:q148} while the \texttt{q1248} distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using the constant PSD with the loss function $\mathcal{L}_{\mathrm{mean}}$. 
	Dashed lines represent the median of the distributions, which has decreased by 10.4\%.}
	\label{fig:q148_q1248_compare}
\end{figure}

%\kw{Which plot is this paragraph referring to?}
Applying the same methods, we will see that the distributions of mismatches
calculated using the {\zdethp} PSD exhibit superior improvement compared to the 
unweighted mismatch.
The shape of the distribution is similar to that shown in Fig. \ref{fig:q148}.
This result is expected, as the IMRPhenomD model was initially constructed and
fitted using the {\zdethp} weighted mismatch. Consequently, the model is
anticipated to closely fit the NR waveforms while incorporating the influence of
the {\zdethp} PSD, rather than a constant PSD.

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248}. We 
specifically choose to use $q=2$ events since we have abundant $q=2$ NR waveforms 
to validate the final result. The new set of coefficients generated from this
optimization process yields only marginal improvements in the newly produced
waveforms, as seen in Fig. \ref{fig:q148_q1248_compare}. The high mismatch tail
of the optimized distribution remains comparable in length and endpoint to the
original distribution, indicating that our procedure is incapable of improving
these waveforms. Similarly, utilizing the {\zdethp} PSD to optimize the loss
function with additional waveforms leads to similar improvement in the resulting
distribution.

@@

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right ($\chi_1>0>\chi_2$) 
regions respond to optimization differently. In the top-left 
region, waveforms generally improve with along optimization. However, mismatches 
in the bottom-right region do not improve after optimization. Most waveforms even 
turned worse after optimization.

\begin{figure}[t]
	\script{ps_q148_qchi.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_qchi.pdf}
	\caption{Parameter space of testing waveforms with $q$ against $\chi_{\mathrm{PN}}$. 
	We show the result from optimizing $\mathcal{L}_{\mathrm{mean}}$ with constant PSD 
	and training waveforms in Tab.~\ref{tab:q148}. Here, the colorbar represents the 
	$\log_{10}$ difference between optimized and original mismatches.}
	\label{fig:ps_q148_qchi}
\end{figure}
\begin{figure}[t]
	\script{ps_q148_chi1chi2.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_chi1chi2.pdf}
	\caption{Parameter space of testing waveforms with $\chi_1$ against $\chi_2$. We
	show the result from optimizing $\mathcal{L}_{\mathrm{mean}}$ with the
	constant PSD and training waveforms listed in Tab.~\ref{tab:q148}.}
	\label{fig:ps_q148}
\end{figure}

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0167 & 8.0 & 0.00     & 0.00     \\ \midrule 
		SXS:BBH:0370 & 1.0 & -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule 
		SXS:BBH:0156 & 1.0 & -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule 
		SXS:BBH:0304 & 1.0 & 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 regions. From top to down are the top-right ($\chi_1,\chi_2>0$), top-left ($\chi_1<0<\chi_2$), bottom-left ($\chi_1,\chi_2<0$) and bottom-right ($\chi_1>0>\chi_2$) regions. Note that for the top-right and bottom-left regions, waveforms are chosen to have $\chi_1\approx\chi_2$, while the training waveforms for the other two regions are chosen to have $\chi_1\approx-\chi_2$.}
	\label{tab:quadrants}
\end{table}

@@

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left and bottom-right regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ as the loss function. Waveforms in the top-left region generally improves while waveforms in the bottom-right region worsened, as indicated by the median (dashed lines).}
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently with waveforms listed in Tab.~\ref{tab:quadrants}. Colorbar represents the $\log_{10}$ difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

In this paper, we have presented a systematic method to recalibrate GW models.
This method utilizes {\jax}'s automatic differentiation to apply
derivative-based optimization to recalibrate GW models jointly. Using the new
implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
mismatches have a slightly better improvement. In contrast, different types of
loss function result in significantly different final mismatch distributions. As 
seen in Fig.~\ref{fig:q148}, $\mathcal{L}_{\mathrm{mean}}$ outperforms 
$\mathcal{L}_{\mathrm{fl}}$. By increasing the number of training waveforms, we see a 
slight improvement increase in Fig.~\ref{fig:q148_q1248_compare}.
